\chapter{Conclusion and Future Work}
\label{ch:conclusion}

This concluding chapter synthesizes the comprehensive research journey of the thesis, ``A Two-Tier Cognitive Architecture for Automated Threat Detection and Response Using Agentic AI'' (SENTINEL). The chapter encapsulates the scientifically proven results mapped back to the core research questions, critically self-assesses the architectural and practical limitations within the scope of a master's thesis, and outlines future research directions for securing modern digital infrastructures using local, context-grounded Agentic AI.

\section{Synthesis of Achieved Research Results}
The synthesis of the achieved research results confirms the successful resolution of the three core research questions proposed at the outset of this study.

The first question (RQ1) concerned stateless wire-speed filtration. The design and implementation of Welford's online statistical algorithm at the transport tier (Tier-1) successfully addresses the challenges of high-volume telemetry processing and alert fatigue. By maintaining a rolling session baseline and calculating Z-Scores in $O(1)$ time and $O(1)$ memory complexity, Tier-1 resolves the overwhelming majority of benign traffic deterministically at wire-speed and escalates only the ambiguous minority to the cognitive layer. This gating is precisely what renders local LLM reasoning operationally feasible: the deterministic tier reaches a verdict in roughly $0.6$\,ms against a multi-second LLM inference, producing the measured $82.97\%$ reduction in mean end-to-end latency relative to an LLM-only pipeline reported in Chapter~\ref{ch:experiments_evaluation}. Furthermore, because Tier-1 maps active session baselines on a per-IP basis, it preserves the temporal event chains of low-and-slow Advanced Persistent Threats (APTs) without dropping the sequential context necessary for long-term correlation.

The second question (RQ2) addressed secure LLM triage and latency mitigation. The integration of a local, quantized Large Language Model (\textit{Gemma-2-9B-IT} Q6\_K) through a LangGraph-driven stateful FSM solves the dual challenges of operational latency and cognitive vulnerability. While the direct ingestion of raw logs by an LLM is bottlenecked by the quadratic complexity of self-attention, the SENTINEL architecture mitigates this by routing only pre-filtered anomalies and by utilizing an in-memory Semantic Cache (an LRU structure keyed on SHA-256 digests of event templates). This cache resolves recurring alerts in microseconds, whereas uncached anomalies average an inference latency of $4$--$6$ seconds per batch on consumer-grade hardware. On the security front, the Delimited Data Encapsulation layer (incorporating cryptographically generated dynamic hexadecimal nonces) together with the Output Sanitization filters proved highly effective at blocking log-substrate prompt injections (S1--S4) and delimiter smuggling, preventing malicious payloads from hijacking the agent's reasoning context.

The third question (RQ3) examined context grounding, memory, and statistical validation. The combination of the Dual-RAG retrieval system and the persistent Threat Memory SQLite database effectively grounds the agent's decisions and minimizes factual hallucination. Utilizing FAISS dense vector search and BM25 sparse keyword search merged via Reciprocal Rank Fusion (RRF, $k=60$), the RAG engine retrieves the relevant MITRE ATT\&CK techniques and NIST playbooks. This grounding was validated through a cross-family LLM-as-a-Judge evaluation, in which RAGAS-style metrics scored Faithfulness at $4.00$, Answer Relevancy at $4.63 \pm 0.80$, Context Recall at $3.99 \pm 0.07$, and Context Precision at $3.00 \pm 0.10$ on a $1$--$5$ scale. Long-term APT correlation is achieved through the Threat Memory schemas (\texttt{ip\_reputation}, \texttt{known\_entities}, \texttt{threat\_events}, and \texttt{apt\_indicators}), which escalate alert levels when a source IP exhibits suspicious activity spanning multiple days or kill-chain phases. Non-parametric statistical tests mathematically reinforce these findings: McNemar's test yielded a p-value of $1.0$ (indicating that Tier-2 adds explanatory power without altering the base F1 classification), and the Mann-Whitney U test yielded $U=27053.5$ with $p = 2.84 \times 10^{-17}$, confirming that the speedup achieved by the two-tier layout is statistically significant.

\section{Scientific and Practical Contributions}
This thesis delivers both scientific and practical contributions to the field of cybersecurity.

\subsection{Scientific Contributions}
The primary theoretical and scientific advancements of this work are threefold. The first is the formalization of a dual-tier cognitive security architecture that decouples transport-layer high-velocity data filtration (Tier-1) from cognitive-layer stateful reasoning (Tier-2), establishing a blueprint for combining deterministic mathematics with non-deterministic language models. The second is the development of a cryptographic isolation model for prompt templates, Delimited Data Encapsulation, that resolves the vulnerability of LLMs to untrusted log substrates without relying on fragile, compute-intensive classifier guardrails. The third is the introduction of a non-parametric evaluation framework for local cybersecurity agents, which employs cross-family LLM judges and RAGAS metrics to measure cognitive alignment, reasoning quality, and contextual grounding.

\subsection{Practical Contributions}
The empirical and engineering contributions of this work are likewise threefold. The first is a fully open-source, containerized prototype (SENTINEL), written in Python and orchestrated via Docker Compose, that integrates \texttt{llama.cpp} and \texttt{LangGraph} for rapid enterprise testing. The second is the validation of local air-gapped threat triage on consumer-grade hardware (an Intel Core i7-14700KF CPU, 32\,GB of DDR5 RAM, and a single NVIDIA RTX 4060 Ti GPU) ensuring data sovereignty and compliance with Zero-Trust guidelines for resource-constrained organizations. The third is an administrative dashboard (Streamlit UI) that bridges automation with human oversight through an interactive Human-in-the-Loop (HITL) queue and secures log files against database tampering by means of blockchain-like HMAC chaining.

\section{Existing Limitations}
Despite the system's demonstrated capabilities, several technical limitations define its operational boundaries. In terms of telemetry and format constraints, the implementation was evaluated primarily on network-flow telemetry and HTTP payload metadata; it lacks ready-to-use parser configurations for diverse unstructured enterprise logs (such as Windows Event Logs, Linux Syslogs, or CloudTrail) which would require additional Drain log-template extraction configurations. With respect to anomaly-queue saturation, although Tier-1 filters the large majority of benign events, a massive network attack such as high-rate scanning or denial of service, by causing a sudden spike in anomalies, would saturate the escalation queue and lead to VRAM constraints and queuing latency unless GPU compute is scaled. A further mitigation trade-off arises because, to prevent an accidental self-denial of service in which the LLM might block critical internal assets (such as Active Directory or database servers), the architecture relies on a human-in-the-loop mechanism to execute firewall-blocking actions; while this ensures safety, it limits the speed of autonomous real-time mitigation. With regard to online semantic poisoning, although vector-database checksums prevent offline file modification, the system remains vulnerable if an adversary can write to the knowledge directories through an authorized channel and thereby inject misleading playbook documents. Finally, the evaluation itself is bounded in scope and statistical power and should be read as indicative rather than exhaustive: the zero-day capability is validated through a controlled statistical proxy (benign flows driven to graded Welford deviations, whose detection boundary is mapped at $\approx 4\sigma$) rather than live malware captures; the APT, adversarial, and reasoning-quality studies rest on modest samples (three campaigns, reported with a Wilson $95\%$ confidence interval of $[0.44, 1.00]$ and a zero-false-firing negative control, together with 120 payloads and a 300-event stratified subset); the full six-configuration ablation (A--F) was evaluated on both an attack-dominated subset and a balanced $150/150$ subset, the latter isolating the Welford gate's precision and latency contribution from the pure-LLM baseline, yet both rest on a single quantized model on a single host, so the absolute figures may shift under other models, hardware, or traffic mixes. The adversarial suite in particular is author-curated, non-adaptive, and aligned to the semantic-downgrade threat model that the tier-consensus guard is designed to neutralize, so it should be read as a proof-of-concept demonstration of the layered mechanism rather than a certification of robustness against an adaptive attacker; evaluation against public injection benchmarks and adaptive attacks is a priority for future work.

\section{Future Development Directions}
Several future development directions emerge from these limitations to guide subsequent research. A first direction is the adoption of multi-agent systems (MAS), decomposing the monolithic LangGraph agent into a cooperative network of specialized sub-agents (such as an IOC-extraction agent, a playbook-correlation agent, a firewall-rule-compiler agent, and a forensic-reporter agent) in order to enhance reasoning quality. A second direction is privacy-preserving federated learning, extending the architecture to a decentralized model in which multiple distributed SENTINEL nodes share locally trained model weights or reputation signals without exchanging raw, privacy-violating network-log payloads. A third direction is active reinforcement learning, integrating a local reinforcement-learning loop in which the agent updates its firewall-rule-generation parameters according to the explicit approvals or rejections issued by SOC analysts in the Streamlit HITL queue. A fourth direction is adversarial self-hardening, deploying a dedicated adversary-simulation agent inside the container network to probe SENTINEL automatically with prompt injections and evasion attacks, allowing the system to test and reinforce its own guardrail tokens dynamically. A fifth direction is TTP-based campaign attribution. The present system detects the \emph{existence} of a multi-day, multi-phase APT campaign in an emergent, unsupervised manner, through multi-day source correlation in the Threat Memory, but deliberately does not name the specific intrusion set responsible. A natural extension is to layer a hypothesis-generating attribution module on top of the structured technique identifiers already produced by the ATT\&CK mapping layer: each observed per-source technique chain would be compared against the published technique profiles of MITRE ATT\&CK Groups using a weighted set-similarity measure, for example a Jaccard or TF-IDF-weighted overlap that up-weights rare techniques, yielding a ranked list of candidate campaigns, each accompanied by a similarity confidence. Because such an extension reuses the existing structured technique output and the per-source kill-chain already maintained in Threat Memory, it would require principally the ingestion of ATT\&CK Group profiles into the knowledge base. It is essential, however, that this output be presented strictly as a ranked hypothesis for analyst review rather than as a definitive verdict: real-world attribution depends on intelligence far beyond shared tradecraft (infrastructure, timing, linguistic artifacts, geopolitical context, and the ever-present possibility of deliberate false-flag mimicry) and because techniques are frequently shared across groups, a purely TTP-driven match can only ever constitute supporting evidence, never proof of authorship.

A further direction is the \emph{production-driven maturation} of the two foundational capabilities that this thesis validates only as a proof-of-concept: signature-less anomaly escalation and emergent multi-day campaign correlation, whose ultimate scope is best fixed by the specific enterprise environment in which SENTINEL is deployed. The Tier-1 zero-day capability is presently a univariate, per-feature Welford Z-score whose detection boundary is characterized at $\approx 4\sigma$; a production deployment tuned to a given organization's traffic profile would extend it toward multivariate or lightweight learned anomaly models and validate it against live malware captures and heterogeneous telemetry rather than the controlled statistical proxies used here. Symmetrically, the multi-day APT correlation is presently keyed on the source IP, which an adversary defeats through IP rotation, proxying, or botnet fan-out; the natural evolution, complementary to the TTP-attribution module above, re-keys correlation on stable techniques and entities so that a campaign remains linkable across changing infrastructure, and enlarges the evaluation beyond the three-campaign corpus reported here. Because both capabilities already reside inside the deployed pipeline (the Welford filter at Tier-1 and the persistent Threat Memory at Tier-2), these are incremental refinements demanded by real operational needs rather than new subsystems.

A methodological direction complements this hardening: a verdict-gated, temporally partitioned re-evaluation of the APT capability. As Chapter~\ref{ch:experiments_evaluation} states plainly, the current APT experiment records an event into the Threat Memory by its DAPT2020 ground-truth annotation, so it validates the multi-day correlation logic rather than the pipeline's own end-to-end detection. A faithful evaluation would instead gate recording on the live two-tier verdict, writing only events that Tier-1 or Tier-2 actually flags, and would partition traffic by \emph{real capture time} rather than at random, since a random split would fabricate the very multi-day structure that defines a campaign. The DAPT2020 corpus is already natively multi-day and CSE-CIC-IDS2018 ships per-date captures, so authentic day boundaries are available without synthetic partitioning. Each simulated day would be padded with a production-scale benign background so that the campaign signal must survive realistic noise, with the $O(1)$ Tier-1 filter run at full volume while the language-model tier is invoked only on the bounded escalated subset. The metrics of interest would then be honest and end-to-end: the true APT recall once low-signal events are subject to the filter, the false-APT rate per simulated day, and the Tier-1 benign-filtration ratio at scale. Such an evaluation is expected to report a lower but more truthful recall and to expose false-APT firings from benign addresses that trip the filter on two separate days, thereby supplying the empirical case for re-keying correlation on techniques and entities rather than on the source IP, exactly as argued above.

\subsection{Horizontal Scalability Roadmap}
A sixth, cross-cutting direction concerns horizontal scalability. The present prototype is deliberately deployed on a single host via Docker Compose and therefore inherits single-node bottlenecks that the two limitations discussed above, anomaly-queue saturation and single-GPU compute, make explicit. A production evolution would follow the established distributed-systems playbook along five coordinated axes, chosen so that the core two-tier cognitive contract remains unchanged.

The first axis is \textit{sharding the stateful Tier-1 filter}. Tier-1 currently holds a per-source-IP Welford baseline in the memory of a single process, which makes it a stateful service that cannot yet scale out. Because each source IP's temporal state is independent of every other, the filter can be partitioned across a fleet of nodes by applying \emph{consistent hashing} to the source-IP key: every IP is deterministically routed to the node that owns its baseline, so per-IP baselines and low-and-slow APT event chains are preserved intact, filter capacity grows linearly with the number of nodes, and the addition or removal of a node re-balances only $k/n$ keys rather than the entire keyspace. This directly converts the sole stateful component into a horizontally scalable tier.

The second axis is \textit{a partitioned, replicated ingestion log}. The single Redis Streams broker used in the prototype is a single point of failure; it would be replaced by a partitioned and replicated commit log (Redis Cluster or Apache Kafka) whose partition key is aligned with the Tier-1 shard key, so that all events from a given source consistently reach the filter node that owns its baseline. The third axis is \textit{idempotent consumption}: because the existing consumer group already guarantees at-least-once delivery, an idempotency layer keyed on a deterministic event identifier would be added so that a redelivery after a crash cannot double-execute a blocking action.

The fourth axis is \textit{an elastic cognitive worker pool}. Tier-2 throughput is presently capped by a single GPU; it would be scaled as a pool of stateless LLM-inference workers that pull from the escalation queue, with the worker count autoscaled against observed queue depth. This directly mitigates the anomaly-queue-saturation limitation (a volumetric burst grows the queue, which in turn provisions additional inference capacity) while the deterministic Tier-1 continues to protect the network independently during inference back-pressure. The fifth axis is \textit{a replicated persistence tier}. The SQLite Threat Memory and audit stores would migrate to a fault-tolerant data layer whose consistency model is chosen per workload: a quorum-based key-value store (a Dynamo- or Cassandra-style design with a tunable $W+R>N$ quorum) suits the eventual-consistency-tolerant IP-reputation and APT-correlation workload, whereas the tamper-evident HMAC audit chain and the human-approved rule store retain strong consistency on a replicated relational engine.

Fronting a stateless dashboard tier with a load balancer and placing an ingestion-side rate limiter ahead of the broker to absorb volumetric floods complete the picture. Taken together, these five axes chart a concrete migration from a validated single-host research prototype to a fault-tolerant, horizontally scalable deployment, addressing the saturation and single-GPU limitations named above, without altering the two-tier cognitive architecture that this thesis validates.

Particularly, to ready the system for enterprise-scale deployment at organizations like FPT Software, the following four radical optimization directions are proposed:
\begin{itemize}
    \item \textbf{Complete Microservice Decoupling:} Instead of Tier-1 and Tier-2 sharing a single Python process (which is bottlenecked by the Global Interpreter Lock - GIL), Tier-2 will be extracted into an independent service communicating with Tier-1 entirely via a Message Broker. This allows Tier-2 to scale out across multiple physical servers limitlessly.
    \item \textbf{Tier-1 Performance Optimization via Golang/Rust:} Rewriting the \texttt{RuleEngine} component of Tier-1 using compiled systems programming languages (such as Golang or Rust) instead of Python to eliminate JSON deserialization overhead and maximize the throughput of Z-Score and WAF Regex calculations, enabling the system to smoothly handle millions of events per second.
    \item \textbf{Adopting Apache Kafka for Persistence:} Replacing Redis Streams with Apache Kafka. While Redis stores data in RAM, posing an Out-of-Memory (OOM) risk if Tier-2 is congested, Kafka writes data to disk sequentially. This ensures that no matter how massive the network traffic burst is, it is safely buffered without crashing the firewall.
    \item \textbf{Utilizing vLLM / TGI for Continuous Batching:} Replacing \texttt{llama.cpp} with \texttt{vLLM} or \texttt{Text Generation Inference (TGI)}. Overcoming the round-robin processing bottleneck of \texttt{llama.cpp}, the PagedAttention technique of \texttt{vLLM} allows batching dozens of network log queries to be processed simultaneously on the GPU in a single clock cycle, multiplying the response speed of Tier-2.
    \item \textbf{Raw Log to APT Inference:} Instead of relying on simulated dataset labels (e.g., \texttt{apt\_phase} from DAPT2020), the Production system will automatically record any IP with a high Z-Score or minor rule violations at Tier-1 into the Threat Memory alongside its timestamp. When an IP accumulates sufficient violations spanning across 2-3 days, the system will automatically trigger an APT Emergent flag and force an ESCALATE action to the AI agent for comprehensive analysis.
\end{itemize}

\vspace{1cm}
\begin{flushright}
    \textit{Closing a modest endeavor, yet opening profound aspirations in the pursuit of securing the digital realm through the power of Agentic AI.}
\end{flushright}
