\chapter{Introduction}
\label{ch:introduction}

This chapter establishes the research framework for SENTINEL, a two-tier cognitive architecture that automates threat detection and incident response in modern Security Operations Centers (SOCs). By analysing alert fatigue, deterministic rule blindness, and the computational constraints of Large Language Models (LLMs), it details the design and validation of a secure hybrid detection pipeline, delineating the objectives, research questions, scope, and quantitative methodology.

\section{Background and Motivation}
The transition towards cloud-native architectures in modern enterprise environments has substantially broadened the potential attack surface. Threat actors now employ a spectrum of techniques ranging from ubiquitous malware to highly orchestrated, multi-stage operations~\cite{mlids2016}. Despite this diversity, defenders consistently face a unified challenge: the overwhelming volume of security telemetry. Security Operations Centers (SOCs) centralise data from Next-Generation Firewalls (NGFWs), Intrusion Detection/Prevention Systems (IDS/IPS), and Endpoint Detection and Response (EDR) solutions into Security Information and Event Management (SIEM) platforms to facilitate normalisation and correlation~\cite{siem2021}. However, conventional front-line mechanisms, such as wire-speed IDS/IPS, remain inherently deterministic. While effective at volume reduction, they lack the capacity to reason about ambiguous signals or provide contextual explanations. Consequently, the sheer velocity of data engenders a ``Big Data Paradox''~\cite{bigdata2019}, disproportionately shifting the cognitive burden of contextual judgment onto human analysts. This structural imbalance culminates in \emph{alert fatigue}, wherein operators are inundated with high volumes of low-fidelity, predominantly false-positive notifications~\cite{alertfatigue2022}, thereby elevating the probability that critical intrusions are inadvertently ignored.

This operational bottleneck is exacerbated by two fundamental gaps. Primarily, traditional signature-based filtration systems~\cite{mlids2016} exhibit structural blindness toward novel attack vectors, such as zero-day exploits and polymorphic variants, which routinely circumvent static rules through trivial payload obfuscation. This highlights an urgent requirement for signature-independent, statistical anomaly detection. Furthermore, economic constraints often dictate log sampling, which inadvertently disrupts the temporal continuity requisite for reconstructing low-and-slow Advanced Persistent Threat (APT) campaigns~\cite{apt_stealth2021}. Conventional Security Orchestration, Automation, and Response (SOAR) platforms fail to bridge this reasoning deficit, as they rely on rigid, pre-defined playbooks incapable of adaptive reasoning or explaining unscripted anomalies~\cite{soar2025}.

Large Language Models (LLMs) offer the sophisticated contextual reasoning necessary to address these shortcomings; nevertheless, their direct deployment within live SOC environments introduces a multifaceted integration paradox. The quadratic computational complexity inherent to the self-attention mechanism~\cite{vaswani2017} renders their inference latency prohibitive for high-velocity telemetry streams. Additionally, the natural-language nature of LLM interfaces creates severe vulnerabilities to log-substrate prompt injections. In such scenarios, adversarial payloads embedded within seemingly innocuous fields (e.g., HTTP User-Agent) can hijack the model's cognitive processes unless rigorous isolation is enforced~\cite{promptinjection2023}. Finally, stringent data sovereignty and Zero-Trust mandates~\cite{nist80207} often prohibit the exfiltration of sensitive telemetry to external cloud services, thereby necessitating that any reasoning engine operate strictly on-premises. To navigate these constraints, this research proposes SENTINEL: a cognitive, dual-tier architecture. It comprises a stateless Tier-1 statistical filter operating at wire-speed, augmented by a learned machine learning (ML) gateway that autonomously resolves the confident majority of escalations. This is seamlessly integrated with a secure, locally hosted Tier-2 Agentic AI tier dedicated to cognitively isolated, high-level reasoning. As an emergent consequence of its stateful orchestration, the design inherently preserves temporal event chains and detects signature-less outliers, establishing a robust foundation for next-generation automated defence.

\section{Research Objectives}
The overarching objective of this thesis is to design, implement, and quantitatively evaluate the SENTINEL architecture, a two-tier cognitive system that synergises low-latency statistical filtration with secure, context-aware agentic reasoning. To realise this, the research pursues three specific objectives:

First, to engineer a dual-stage filtration pipeline (\textbf{Tier 1}) that mitigates alert fatigue at the ingestion layer. This involves implementing a wire-speed deterministic filter leveraging Welford's online algorithm for $O(1)$ rolling variance and Z-score computation, sequentially coupled with a learned Machine Learning (ML) gateway (LightGBM). The primary goal is to maximise the autonomous resolution of confident escalations, thereby drastically reducing the invocation frequency of the computationally expensive LLM inference engine while maintaining measurable detection accuracy.

Second, to architect secure, structural encapsulation mechanisms for the integration of a local, quantised LLM (\textbf{Tier 2}). This necessitates the design of deterministic, construction-based defences---specifically Delimited Data Encapsulation utilizing cryptographic nonces (e.g., \texttt{<<<DATA\_HASH...>>>}), a bi-tier consensus guardrail, output sanitisation, and an HMAC-chained audit trail. The objective is to neutralise adversarial vectors, such as log-substrate prompt injections, through architectural guarantees rather than relying on brittle, training-dependent semantic classifiers.

Third, to construct a stateful cognitive orchestration framework leveraging LangGraph, integrating a persistent Threat Memory and a dual retrieval-augmented grounding subsystem. By fusing dense semantic search with sparse keyword matching (MITRE ATT\&CK and NIST SP~800-61r2) via Reciprocal Rank Fusion, this objective aims to robustly ground the agent's forensic attributions and justifications, whilst enabling the emergent correlation of multi-day campaigns derived from accumulated state.

\section{Research Questions}
\label{sec:rq}
The empirical validation of the proposed architecture is directed by three central research questions, each corresponding directly to the aforementioned objectives:

\textbf{RQ1 (Filtration Economics):} How can a hybrid pipeline---comprising an $O(1)$ deterministic filter based on online statistical baselines and a subsequent learned ML gateway---effectively absorb high-velocity SOC telemetry prior to invoking multi-second local LLM inference, and what is the quantifiable trade-off regarding detection accuracy? This question emphasizes the architectural value of operationalizing a local LLM by offloading the majority of processing to faster, specialized components.

\textbf{RQ2 (Architectural Defence Boundaries):} Within the architecture, where is defence against log-substrate injection most effectively situated, and how do the capabilities of deterministic pattern filtration contrast with Delimited Data Encapsulation and two-tier consensus guardrails? Specifically, which adversarial vectors are neutralised by each respective layer, and how does this structural approach secure the cognitive tier against downgrade attacks?

\textbf{RQ3 (The Efficacy of Statefulness):} What specific capabilities are enabled by transitioning from rigid Runbook Automation to a stateful agentic framework equipped with persistent Threat Memory and dual-retrieval mechanisms? This is measured by evaluating (a) the emergent capacity to correlate multi-day campaigns without prior explicit temporal rules, and (b) the precision of MITRE ATT\&CK technique attribution alongside the fidelity of the generated justifications.

\begin{quote}\small
\textbf{Note on Instrumentation.} RQ3 intentionally refrains from employing binary classification accuracy as a metric for assessing the contribution of knowledge retrieval. The retrieved context fundamentally influences the \emph{attribution} and \emph{justification} of a verdict rather than the binary determination of maliciousness. Applying binary accuracy in this context would constitute a methodological mismatch. Similarly, this thesis does not claim the LLM tier functions as a primary classifier; empirical evidence positions it as a coverage-biased safety net for complex reasoning, a limitation explicitly acknowledged in Chapter~\ref{ch:conclusion}.
\end{quote}

\section{Scope and Object of Study}
\textbf{Object of Study:} The primary subject of this research is the two-tier cognitive architecture itself. This encompasses the precise mechanisms through which a stateless, wire-speed Tier-1 statistical filter and a learned gateway are orchestrated alongside a stateful, locally hosted Tier-2 agentic framework to automate SOC detection and incident response. This includes high-velocity OSI Layer 3--4 flow telemetry (specifically, the NetFlow five-tuple and CICFlowMeter features, utilizing PCAP metadata in lieu of Deep Packet Inspection); stateful agent orchestration via LangGraph; the execution of air-gapped inference using local quantised LLMs; and the mitigation of adversarial vectors such as log-substrate injection, delimiter smuggling, and knowledge-base poisoning.

\textbf{Scope:} The empirical evaluation is strictly constrained to the CSE-CIC-IDS2018 dataset~\cite{ids2018} for broad attack classification, and the DAPT2020 dataset~\cite{dapt2020} for multi-stage APT correlation. Zero-day threats are simulated by injecting extreme Welford statistical deviations into verified benign flows. The reasoning engine is standardized on a locally hosted \textit{Gemma-2-9B-IT} model, quantised to the Q6\_K format via \texttt{llama.cpp}, executing on a single consumer-grade GPU to strictly enforce data sovereignty. Furthermore, autonomous response capabilities are limited to the generation of firewall policies, forensic reporting, and structural attributions; automated physical network disconnection is explicitly excluded from the current scope.

\section{Research Methodology}
This thesis adopts a Design Science Research (DSR) methodology, underpinned by deductive and quantitative reasoning. It generates novel knowledge by constructing a functional artifact (SENTINEL) and rigorously evaluating its performance against real-world operational challenges, ensuring all theoretical claims are empirically substantiated. The reasoning framework operates across three dimensions: statistical (delineating normal traffic baselines from anomalies), logical (formalising the agent's decision-making state-graph), and adversarial (stress-testing systemic resilience against manipulation).

Practically, the architecture is implemented as a functional two-tier prototype. It undergoes end-to-end evaluation by replaying archived datasets and simulated live traffic, specifically introducing adversarial payloads to assess robustness. The system is evaluated across a comprehensive five-dimensional framework: Accuracy, Performance, Security, Explainability, and Integrity. Controlled ablation studies are employed to systematically isolate the contribution of individual components, while non-parametric statistical tests (e.g., McNemar's and Mann--Whitney U) confirm the statistical significance of observed architectural variations. The evaluation synthesises quantitative metrics (e.g., latency, F1-scores) with qualitative outputs (the model's forensic justifications). To prevent self-aggrandisement bias, qualitative reasoning is objectively scored utilizing an automated RAGAS-inspired framework~\cite{ragas2023} administered by an independent, cross-family LLM judge, ensuring a rigorous, comparative assessment.

\section{Contributions of the Thesis}
\label{sec:contributions}
The contributions of this research are tri-fold, each strictly delineated by the supporting empirical evidence.

\textbf{Architecturally}, the thesis validates a hybrid two-tier pipeline that successfully resolves the latency bottlenecks associated with locally hosted, multi-second LLMs in operational SOCs. The primary contribution is the \emph{compositional integration}: a stateless $O(1)$ statistical filter combined with a learned LightGBM gateway that autonomously sheds $83.8\%$ of confident escalations at wire-speed. Furthermore, an in-memory Semantic Cache (Tier-1.75) resolves recurring alerts in approximately $1$\,ms. This ensures the network is protected deterministically, reserving the computationally intensive cognitive tier exclusively for the ambiguous minority of anomalous events.

\textbf{In AI Security}, the research contributes a deterministic, structurally robust defence mechanism against log-substrate prompt injections. By employing Delimited Data Encapsulation utilizing cryptographic nonces (e.g., \texttt{<<<DATA\_HASH...>>>}), a deterministic Tier-1/Tier-2 consensus guardrail, and an HMAC-chained audit trail, the system provides security guarantees that hold by architectural construction rather than probabilistic training. This ensures that the fixed-rule consensus guard cannot be persuasively bypassed, effectively neutralizing both syntactic and semantic downgrade attempts.

\textbf{In Orchestration}, the architecture supersedes rigid Runbook Automation with a stateful, acyclic, and tool-free state-graph orchestrated on a single local quantised model. This yields a response tier that is highly adaptive, yet strictly reproducible and forensically auditable, entirely within the constraints of strict data sovereignty. Additional foundational capabilities enabled by this architecture---namely, signature-less anomaly escalation via statistical baselining and emergent multi-day APT correlation from persistent state---are presented as bounded proofs of concept, laying the groundwork for future production-scale maturation.

\section{Structure of the Thesis}
Chapter~\ref{ch:theoretical_background} establishes the theoretical foundations, including LLM quantization, state-graph models, hybrid RAG, online statistics, and reviews related agentic-security literature. Chapter~\ref{ch:system_design_implementation} details the technical system design: the two-tier filter, the ML gateway, the stateful agentic Finite State Machine (FSM), hybrid retrieval, structural guardrails, the dashboard, and deployment configurations. Chapter~\ref{ch:experiments_evaluation} presents the empirical experiments, controlled ablation studies, and non-parametric statistical validation. Finally, Chapter~\ref{ch:conclusion} synthesises the findings, acknowledges existing limitations, and outlines future research trajectories.
