\chapter{Conclusion and Future Work}
\label{ch:conclusion}

This chapter synthesises the research journey of the thesis, ``A Two-Tier Cognitive Architecture for Automated Threat Detection and Response Using Agentic AI'' (SENTINEL). It maps the proven results back to the research questions, self-assesses the architectural and practical limitations, and outlines future directions for securing modern infrastructures with local, context-grounded Agentic AI.

\section{Synthesis of Achieved Research Results}
The results confirm the successful resolution of the three research questions.

\textbf{RQ1 (the economics of the filtration tier).} The two-stage filtration chain absorbs the overwhelming majority of traffic before multi-second reasoning is required. Maintaining a rolling per-IP baseline and computing Z-scores in $O(1)$ time and memory, Tier-1 deterministically resolves the benign majority; a \emph{learned LightGBM gateway} then adjudicates $83.8\%$ of the remaining escalations without an LLM call, at $98.82\%$ precision on the portion it intercepts. Each fast-path decision completes in well under a millisecond (the gateway adds $\approx\!0.35$\,ms) against a multi-second inference---producing the measured $82.97\%$ reduction in mean end-to-end latency relative to an LLM-only pipeline. \emph{The trade-off} is explicitly stated: the gateway attains a recall of $0.7551$ on the test set, meaning the traffic portion it absorbs trades coverage for throughput---this is the central trade-off posed by RQ1, not a side effect.

\textbf{RQ2 (secure, low-latency triage).} A local quantised \textit{Gemma-2-9B-IT} Q6\_K model driven by a LangGraph FSM solves operational latency and cognitive vulnerability. Only pre-filtered anomalies reach the model, and an in-memory Tier-1.75 Semantic Cache (LRU keyed on SHA-256 template digests) resolves recurring alerts in $\approx\!1$\,ms while uncached anomalies average $4$--$6$\,s per batch, supplemented by prompt KV-Caching that further reduces local TTFT. On security, Delimited Data Encapsulation (\texttt{<<<DATA\_HASH...>>>} nonces) plus output sanitisation block log-substrate prompt injections (S1--S4) and delimiter smuggling; because the tier-consensus guard is deterministic, the full pipeline resists every downgrade attempt in the curated suite.

\textbf{RQ3 (what capabilities does state afford).} Accumulated state yields a capability that a stateless tier cannot possess. \emph{(a)~Emergent multi-day correlation:} The persistent Threat Memory detects $3/3$ multi-day APT campaigns (recall $1.00$), and---more importantly---a \emph{negative control} confirms a specificity of $1.00$: four benign IPs that also operated across $\ge 2$ distinct days triggered zero APT flags. Recall alone is meaningless here (a naive counter shouting ``APT'' for every multi-day IP would also score $3/3$); it is the negative control that proves the system can \emph{discriminate}. \emph{(b)~Attribution and justification:} FAISS dense and BM25 sparse retrieval fused by RRF ($k=60$) achieve a $64.0\%$ technique exact-match on the application-attack class---against a knowledge base coverage ceiling of $80.0\%$---and a cross-family LLM-as-Judge (Meta Llama-3-8B judging Gemma-2-9B, $n=908$) scored Faithfulness $3.30$, Answer Relevancy $3.56\pm1.25$, Context Recall $3.27\pm0.95$, and Context Precision $2.25\pm0.57$ (overall $3.1/5$). Retrieval precision is the weak link and is noted in Section~\ref{sec:limitations}. Regarding statistical validation, the Mann--Whitney~U test confirms the two-tier latency separation is highly significant ($p<0.001$); McNemar's test on the attack-heavy subset yields $p=1.0$, and this thesis \emph{does not} interpret this as a detection superiority of Tier-2---see the discussion on measurement instrument limitations in Section~\ref{sec:limitations}.

\section{Scientific and Practical Contributions}
\textbf{Scientific.} (1)~A dual-tier cognitive security architecture that decouples transport-layer high-velocity filtration (Tier-1, deterministic statistics plus a learned load-shedding gateway) from cognitive-layer stateful reasoning (Tier-2), a blueprint for combining deterministic mathematics with non-deterministic language models. (2)~A cryptographic isolation model, Delimited Data Encapsulation, that resolves LLM vulnerability to untrusted log substrates without fragile, compute-intensive classifier guardrails. (3)~A non-parametric evaluation framework for local security agents using cross-family LLM judges and RAGAS metrics.

\textbf{Practical.} (1)~A fully open-source, containerised prototype in Python orchestrated via Docker Compose, integrating \texttt{llama.cpp} and \texttt{LangGraph}. (2)~Validated local air-gapped triage on consumer hardware (i7-14700KF, 32\,GB RAM, one RTX~4060~Ti), with a LightGBM gateway that offloads $83.8\%$ of escalations at $98.82\%$ precision to keep local inference feasible. (3)~A Streamlit dashboard bridging automation with a Human-in-the-Loop queue, securing logs against tampering via HMAC chaining.

\section{Existing Limitations}
Several limitations bound the operational scope. \emph{Telemetry:} evaluation focused on network-flow and HTTP metadata; diverse unstructured logs (Windows Event, Syslog, CloudTrail) would need further Drain configurations. \emph{Queue saturation:} a high-rate scan or DoS could spike anomalies and saturate the escalation queue under single-GPU compute. \emph{Autonomy trade-off:} to avoid self-denial of service, firewall blocks are human-gated, which limits fully autonomous mitigation speed. \emph{Online poisoning:} vector checksums prevent offline tampering but not writes through an authorised channel. \emph{Statistical power:} the zero-day capability is validated by a controlled Welford proxy ($\approx\!4\sigma$ boundary) rather than live malware; the APT, adversarial, and reasoning studies rest on modest samples (three campaigns, Wilson $95\%$ CI $[0.44,1.00]$ with a zero-false-firing control; a 120-payload suite; a $1{,}250$-sample attack-dominated and a balanced $150/150$ ablation, both on a single model and host). The adversarial suite is author-curated, non-adaptive, and matched to the semantic-downgrade model the consensus guard neutralises, so it is a proof-of-concept, not a robustness certification; evaluation against public injection benchmarks and adaptive attacks is a priority.

\subsection{Limitations of the Measurement Instrument}
\label{sec:limitations}
Two limitations belong to the \emph{measurement}, not the artifact, and are noted here because they strongly bound how the figures should be read.

\emph{First, the resolution of the binary metric.} When collapsing all verdicts into a ``detection or not'' label, the \texttt{ESCALATE} (Tier-1 forwarding) and \texttt{AWAIT\_HITL} (agent deferring to a human) actions are grouped into the same bin as \texttt{BLOCK}. Consequently, a configuration that \emph{defers the decision downstream} is scored identically to one that \emph{decides correctly}, causing the ablation configurations to converge on the same value and rendering the comparison indistinguishable---this is exactly why McNemar yields $p=1.0$ above. Therefore, the thesis \emph{does not} draw conclusions about the per-component detection contribution from the binary metric; claims regarding component contributions rest solely on isolated measurements (ML gateway offload, MITRE attribution, APT negative control, adversarial suite). A metric that \emph{scores based on the final action} against the expected action---distinguishing correct auto-resolutions, incorrect auto-resolutions, and deferrals---is a prerequisite for quantifying per-component contributions and is noted as immediate future work.

\emph{Second, the LLM tier is not a classifier.} When deliberately bypassing the ML gateway and feeding a $95\%$ benign subset into the agent ($n=800$), the LLM tier achieved a recall of $1.00$ but a specificity of $0.00$, with an accuracy of $0.0475$ \emph{exactly equalling} the majority baseline---meaning in that regime, it flagged almost everything. This figure does not describe the true operational path (where the ML gateway pre-filters), but it establishes an honest boundary: the cognitive tier acts as a \emph{coverage-biased safety net} with justification capabilities, not as a benign/malicious discriminator. All statements in the thesis regarding this tier are kept within that boundary.

\section{Future Development Directions}
Several directions follow. A \emph{multi-agent} decomposition would split the monolithic agent into cooperating sub-agents (IOC extraction, playbook correlation, rule compilation, forensic reporting). \emph{Federated learning} would share locally trained weights or reputation signals across nodes without exchanging raw logs. \emph{Reinforcement learning} would update rule-generation parameters from analyst approvals in the HITL queue. \emph{Adversarial self-hardening} would deploy an in-network adversary agent to probe the guardrails automatically. \emph{TTP-based attribution} would layer a hypothesis-generating module on the structured ATT\&CK identifiers, comparing per-source technique chains against MITRE ATT\&CK Group profiles by weighted set-similarity---presented strictly as a ranked hypothesis for analyst review, never a definitive verdict, since techniques are shared across groups and real attribution needs intelligence far beyond tradecraft.

Two foundational capabilities warrant \emph{production-driven maturation}. The Tier-1 zero-day filter is presently a univariate per-feature Welford Z-score ($\approx\!4\sigma$ boundary); production would extend it toward multivariate or lightweight learned models validated against live malware. The multi-day APT correlation is presently keyed on the source IP (defeated by IP rotation or botnet fan-out); re-keying on stable techniques and entities would keep campaigns linkable across changing infrastructure. Methodologically, a faithful APT re-evaluation would gate Threat-Memory recording on the \emph{live} two-tier verdict (not the DAPT2020 annotation) and partition traffic by real capture time, padding each day with production-scale benign background---yielding a lower but truthful end-to-end recall and exposing false-APT firings, thereby motivating the technique/entity re-keying above.

\subsection{Horizontal Scalability Roadmap}
The single-host prototype inherits single-node bottlenecks. A production evolution follows the distributed-systems playbook without altering the two-tier contract: (i)~\emph{shard the stateful Tier-1 filter} by consistent hashing on the source-IP key, so per-IP baselines and APT chains are preserved while capacity grows linearly; (ii)~replace the single Redis broker with a \emph{partitioned, replicated commit log} (Redis Cluster or Kafka) whose partition key aligns with the Tier-1 shard key; (iii)~add \emph{idempotent consumption} keyed on a deterministic event identifier so a redelivery cannot double-execute a block; (iv)~scale Tier-2 as an \emph{elastic pool} of stateless inference workers autoscaled against queue depth, directly mitigating saturation while Tier-1 protects independently; and (v)~migrate the SQLite stores to a \emph{replicated persistence tier} (a quorum key-value store for reputation/APT, strong consistency for the HMAC audit and rule store). A stateless dashboard behind a load balancer and an ingestion-side rate limiter complete the picture. For enterprise scale, five concrete optimisations apply:
\begin{itemize}
    \item \textbf{Microservice decoupling:} extract Tier-2 into an independent service communicating via a message broker, escaping the Python GIL and scaling out across servers.
    \item \textbf{Compiled Tier-1 (Go/Rust):} rewrite the \texttt{RuleEngine} in a systems language to eliminate JSON overhead and sustain millions of events per second.
    \item \textbf{Kafka persistence:} replace Redis Streams with disk-backed Kafka to buffer volumetric bursts without OOM risk.
    \item \textbf{vLLM / TGI batching:} replace \texttt{llama.cpp} with PagedAttention-based continuous batching to adjudicate dozens of escalations per GPU cycle.
    \item \textbf{SQLite \& Redis Stability:} enforce \texttt{synchronous=NORMAL} (strictly avoiding \texttt{WAL} mode) and backpressure via consumer group \texttt{lag} to prevent cross-UID crash loops in Docker.
    \item \textbf{Verdict-driven APT recording:} in production, record any high-Z or rule-tripping IP into Threat Memory with its timestamp, auto-flagging an emergent APT once violations span 2--3 days.
\end{itemize}

\vspace{0.6cm}
\begin{flushright}
    \textit{Closing a modest endeavor, yet opening profound aspirations in the pursuit of securing the digital realm through the power of Agentic AI.}
\end{flushright}
