\chapter{Technical Implementation}
\label{ch:implementation}

% Chapter introduction
This chapter details the software engineering aspects and practical source code implementation of the SENTINEL architecture. Shifting from abstract theory, the content focuses on the project directory organization, core class structures (Classes/Methods), Docker service configurations, and the design of the Graphical User Interface (Streamlit Dashboard) which facilitates Human-in-the-Loop (HITL) interaction.

\section{Technology Stack and Source Code Structure}
The core technology stack chosen for the implementation of SENTINEL consists of Python 3.10+, LangGraph for stateful multi-agent orchestration, LangChain for LLM integrations, FAISS for dense vector search, SQLite for local threat databases and memory, Streamlit for the user interface, Docker for virtualization, and llama.cpp via Llama-cpp-python bindings to support local, quantized inference. The directory structure is organized as a modular tree representation that strictly separates system concerns: the folder \texttt{src/core/} contains the main Tier-1 and Tier-2 logic; \texttt{src/rag/} manages vector database ingestion and search; \texttt{src/security/} implements guardrails and input-output filters; \texttt{src/dashboard/} houses the Streamlit user interface; and \texttt{experiments/} and \texttt{tests/} contain the validation test suites.

\section{Tier-1 Implementation: Filtration Engine \& Welford}
The Tier-1 filtration engine is implemented using two core classes, namely \texttt{RunningStats} and \texttt{SessionBaseline}. The \texttt{RunningStats} class translates Welford's algorithm into Python, computing numerical variance and means on-the-fly. To prevent infinite memory growth from tracking session states of all source IPs, the system hashes IP addresses and incorporates a garbage-collection mechanism via a Time-To-Live (TTL) cache to purge inactive sessions. The main filtering logic is encapsulated within the \texttt{RuleEngine} class, which executes fast pattern matching using compile-time regular expressions and static thresholds to intercept known attack strings.

\section{Tier-2 Implementation: AI Agent and RAG}
The Tier-2 cognitive agent is driven by the \texttt{OpenAILLMClient} class, which wraps the local llama.cpp endpoint. It enforces specific parameters such as setting \texttt{temperature=0.0} to maximize response determinism, bounding the context size to avoid memory exhaustion, and configuring \texttt{n\_gpu\_layers} to offload model computations to the GPU. The workflow orchestration is set up using LangGraph's \texttt{StateGraph} API, which defines the state transitions between the \texttt{analyze\_node}, \texttt{retrieve\_node}, and a conditional router function named \texttt{should\_retrieve} that guides the execution flow. Retrieval logic is handled by the \texttt{DualRetriever} class, which combines the FAISS index for vector search with the \texttt{rank\_bm25} package for keyword matching, merging and sorting their respective outputs using the Reciprocal Rank Fusion algorithm.

\section{Guardrail and HMAC System Implementation}
Defensive guardrails and audit mechanisms are implemented within the \texttt{DelimitedDataEncapsulator} and \texttt{HMACHashing} classes. The \texttt{DelimitedDataEncapsulator} class contains the logic to generate cryptographically secure hexadecimal delimiting tokens using Python's \texttt{os.urandom} and \texttt{hashlib} packages, wrapping untrusted log payloads inside these transient delimiters before passing them to the prompt template. Database integrity is managed by the \texttt{HMACHashing} class, which interacts with the SQLite database (\texttt{threat\_memory.db}). It defines trigger functions that dynamically calculate SHA-256 HMAC hashes, linking each new threat event record directly to the hash of the preceding entry in a blockchain-like design.

\section{SOC Administration Dashboard Implementation (Streamlit UI)}
The Streamlit-based SOC Administration Dashboard follows a reactive model where the Streamlit frontend communicates dynamically with the database and files of the backend to update UI elements. The dashboard integrates three specialized display modules. First, a real-time queue interface visualizes Tier-1 telemetry throughput and Tier-2 analysis verdicts. Second, a threat memory analytics panel plots APT alert propagation graphs across IP-based temporal timelines. Third, an audit trail integrity module verifies the HMAC chain in the database, displaying status indicators where a green light signals a valid chain and a red light warns that logs have been modified. Additionally, the dashboard incorporates a Human-in-the-Loop interface, allowing analysts to manually review and either approve or reject the agent's recommended firewall blocks, updating live firewall configurations on-the-fly.

\section{Containerization and Infrastructure Deployment}
System deployment is virtualized using Docker, specified in a multi-stage \texttt{Dockerfile} designed to minimize the final production image size. The build process compiles llama.cpp with CUDA support to enable GPU-accelerated local inference within the container. Infrastructure services are managed via \texttt{docker-compose.yml}, which links the agent reasoning container with the Streamlit dashboard container, sharing log and database directories using persistent volume mounts.

In summary, Chapter 4 systematizes the software development process of SENTINEL, bridging core algorithms with user interfaces and virtualized infrastructure. Chapter 5 will subsequently establish the testing environment to quantitatively evaluate these components.
