\chapter{Conclusion and Future Work}
\label{ch:conclusion}

This concluding chapter synthesizes the comprehensive research journey of the thesis, ``A Two-Tier Cognitive Architecture for Automated Threat Detection and Response Using Agentic AI'' (SENTINEL). The chapter encapsulates the scientifically proven results mapped back to the core research questions, critically self-assesses the architectural and practical limitations within the scope of a master's thesis, and outlines future research directions for securing modern digital infrastructures using local, context-grounded Agentic AI.

\section{Synthesis of Achieved Research Results}
The synthesis of the achieved research results confirms the successful resolution of the three core research questions (RQ1, RQ2, and RQ3) proposed at the outset of this study:
\begin{enumerate}
    \item \textbf{Resolution of RQ1 (Stateless Wire-Speed Filtration):} The design and implementation of Welford's online statistical algorithm at the Transport Tier (Tier-1) successfully addresses the challenges of high-volume telemetry processing and alert fatigue. By maintaining a rolling session baseline and calculating Z-Scores in $O(1)$ time and $O(1)$ memory complexity, Tier-1 achieved a $99.8\%$ benign log filtration rate on the CSE-CIC-IDS2018 dataset. This massive reduction ensures that only $0.2\%$ of anomalous network events are escalated to the computationally expensive cognitive layer. Furthermore, because Tier-1 maps active session baselines on a per-IP basis, it preserves the temporal event chains of low-and-slow Advanced Persistent Threats (APTs) without dropping the sequential context necessary for long-term correlation.
    \item \textbf{Resolution of RQ2 (Secure LLM Triage and Latency Mitigation):} The integration of a local, quantized Large Language Model (\textit{Gemma-2-9B-IT} Q6\_K) via a LangGraph-driven stateful FSM solves the dual challenges of operational latency and cognitive vulnerability. While direct ingestion of raw logs by an LLM is bottlenecked by the quadratic complexity of self-attention, the SENTINEL architecture mitigates this by routing only pre-filtered anomalies and utilizing an in-memory Semantic Cache (LRU-based with SHA-256 keys of event templates). This cache resolves recurring alerts in microseconds, while uncached anomalies average an inference latency of $4$--$6$ seconds per batch on consumer-grade hardware. On the security front, the Delimited Data Encapsulation layer (incorporating cryptographically generated dynamic hexadecimal nonces) and Output Sanitization filters proved highly effective at blocking log-substrate prompt injections (S1--S4) and delimiter smuggling, preventing malicious payloads from hijacking the agent's reasoning context.
    \item \textbf{Resolution of RQ3 (Context Grounding, Memory, and Statistical Validation):} The combination of the Dual-RAG retrieval system and the persistent Threat Memory SQLite database effectively grounds the agent's decisions and minimizes factual hallucinations. Utilizing FAISS dense vector search and BM25 sparse keyword search merged via Reciprocal Rank Fusion (RRF, $k=60$), the RAG engine retrieves relevant MITRE ATT\&CK techniques and NIST playbooks. This grounding was validated through cross-family LLM-as-a-Judge evaluation, where RAGAS-style metrics scored Faithfulness at $4.00$, Answer Relevancy at $4.63 \pm 0.80$, Context Recall at $3.99 \pm 0.07$, and Context Precision at $3.00 \pm 0.10$ on a $1$--$5$ scale. Long-term APT correlation is achieved through the Threat Memory schemas (\texttt{ip\_reputation}, \texttt{known\_entities}, \texttt{threat\_events}, and \texttt{apt\_indicators}), which successfully escalate alert levels when a source IP shows suspicious activities spanning multiple days or kill-chain phases. Non-parametric statistical tests mathematically reinforce these findings: McNemar's test yielded a p-value of $1.0$ (proving Tier-2 adds explanatory power without altering the base F1 classification), and the Mann-Whitney U test yielded $U=27053.5$ with $p = 2.84 \times 10^{-17}$, confirming the speedup achieved by the two-tier layout is statistically significant.
\end{enumerate}

\section{Scientific and Practical Contributions}
This thesis delivers both scientific and practical contributions to the field of cybersecurity.

\subsection{Scientific Contributions}
The primary theoretical and scientific advancements of this work include:
\begin{itemize}
    \item The formalization of a dual-tier cognitive security architecture that decouples transport-layer high-velocity data filtration (Tier-1) from cognitive-layer stateful reasoning (Tier-2), establishing a blueprint for combining deterministic mathematics with non-deterministic language models.
    \item The development of a cryptographic isolation model for prompt templates (Delimited Data Encapsulation) that resolves the vulnerability of LLMs to untrusted log substrates without relying on fragile, compute-intensive classifier guardrails.
    \item The introduction of a non-parametric evaluation framework for local cybersecurity agents, utilizing cross-family LLM judges and RAGAS metrics to measure cognitive alignment, reasoning quality, and contextual grounding.
\end{itemize}

\subsection{Practical Contributions}
The empirical and engineering contributions of this work include:
\begin{itemize}
    \item A fully open-source, containerized prototype (SENTINEL) written in Python and orchestrated via Docker Compose, integrating \texttt{llama.cpp} and \texttt{LangGraph} for rapid enterprise testing.
    \item Validation of local air-gapped threat triage on consumer-grade hardware (Intel Core i7-14700KF, 32GB DDR5 RAM, and a single NVIDIA RTX 4060 Ti GPU), ensuring data sovereignty and compliance with Zero-Trust guidelines for resource-constrained organizations.
    \item An administrative dashboard (Streamlit UI) that bridges automation with human oversight via an interactive Human-in-the-Loop (HITL) queue and secures log files against database tampering through blockchain-like HMAC chaining.
\end{itemize}

\section{Existing Limitations}
Despite the system's demonstrated capabilities, several technical limitations define its operational boundaries:
\begin{itemize}
    \item \textbf{Telemetry and Format Constraints:} The implementation was evaluated primarily on Network Flow telemetry and HTTP payload metadata. It lacks ready-to-use parser configurations for diverse unstructured enterprise logs (e.g., Windows Event Logs, Linux Syslogs, CloudTrail), which would require additional Drain log-template extraction configurations.
    \item \textbf{Anomaly Queue Saturation:} Although the Welford filter drops $99.8\%$ of benign events, a massive network attack (e.g., high-rate scanning or denial of service) causing a sudden spike in anomalies would saturate the escalation queue, leading to VRAM constraints and queuing latency if GPU compute is not scaled.
    \item \textbf{Mitigation Trade-Off:} To prevent accidental Self-Denial of Service (Self-DoS) where the LLM might block critical internal assets (e.g., Active Directory or database servers), the architecture relies on a Human-in-the-Loop mechanism for executing firewall blocking actions. While this ensures safety, it limits the speed of autonomous real-time mitigation.
    \item \textbf{Online Semantic Poisoning:} While vector database checksums prevent offline file modification, the system remains vulnerable to online semantic poisoning if an adversary can write to the knowledge directories through an authorized channel, injecting misleading playbook documents.
\end{itemize}

\section{Future Development Directions}
Several future development directions emerge from these limitations to guide subsequent research:
\begin{itemize}
    \item \textbf{Multi-Agent Systems (MAS):} Decomposing the monolithic LangGraph agent into a cooperative multi-agent network (comprising specialized sub-agents such as an IOC Extraction Agent, a Playbook Correlation Agent, a Firewall Rule Compiler Agent, and a Forensic Reporter Agent) to enhance reasoning quality.
    \item \textbf{Privacy-Preserving Federated Learning:} Extending the architecture to a decentralized model where multiple distributed SENTINEL nodes share locally trained model weights or reputation anomalies without exchanging raw, privacy-violating network log payloads.
    \item \textbf{Active Reinforcement Learning:} Integrating a local reinforcement learning loop where the agent updates its firewall rule generation parameters based on explicit approvals or rejections from SOC analysts in the Streamlit HITL queue.
    \item \textbf{Adversarial Self-Hardening:} Deploying a dedicated adversary simulation agent inside the container network to automatically probe SENTINEL with prompt injections and evasion attacks, allowing the system to dynamically test and reinforce its own guardrail tokens.
\end{itemize}

\vspace{1cm}
\begin{flushright}
    \textit{Closing a modest endeavor, yet opening profound aspirations in the pursuit of securing the digital realm through the power of Agentic AI.}
\end{flushright}
