\chapter{Experiments and Evaluation}
\label{ch:experiments_evaluation}

% Chapter introduction
This chapter presents the empirical evidence of the thesis, validating the research hypotheses through rigorous quantitative metrics. The SENTINEL system is evaluated against a 5-Dimensional framework (5D Metrics: Accuracy, Performance, Security, Explainability, and Integrity). The content emphasizes the layered Ablation Study design and the application of biostatistical models (McNemar's Test, Mann-Whitney U Test) to prove that the architectural enhancements are statistically significant and mathematically sound.

\section{Experimental Environment and Datasets}
To execute the empirical evaluations, a dedicated local hardware testbed was configured. The hardware setup consists of an Intel Core i9 CPU, 32GB of RAM, and an NVIDIA RTX 4060 Ti 16GB GPU, which was utilized to execute the local quantized \textit{Gemma-2-9B-IT} Q4\_K\_M model. The evaluations rely on two major datasets: the CSE-CIC-IDS2018 dataset, which encompasses diverse network attack typologies—including Brute Force, DoS, and Web Attacks—serving as the baseline for evaluating classification accuracy; and the DAPT2020 dataset, which is employed for testing the system's capacity to correlate multi-stage Advanced Persistent Threat (APT) campaigns over prolonged timelines. To replicate the properties of a live SIEM environment, a Unified Data Stream simulator was scripted to merge and replay the network logs chronologically according to virtual timestamps, feeding them to the detection queues in real time.

\section{5D Evaluation Metrics and Ablation Study Design}
The analytical performance of the system is benchmarked across a five-dimensional evaluation framework. These 5D metrics include Accuracy, measured by Precision, Recall, and F1-Score; Performance, measuring latency and throughput; Security, defined by the adversarial robustness rate; Explainability, evaluated via audit completeness; and Integrity, verified by HMAC validation. To isolate the quantitative impact of each proposed component, an ablation study was designed across six specific configurations. Configuration A acts as the first baseline, relying exclusively on a traditional rule-based signature engine. Configuration B serves as the second baseline, utilizing pure local LLM inference without Welford pre-filtering or RAG context. Configuration C introduces Welford's Tier-1 filter before the LLM. Configuration D adds a single-RAG FAISS vector search to the LLM. Configuration E expands the RAG component to a hybrid FAISS and BM25 configuration. Finally, Configuration F represents the holistic SENTINEL system, combining the Tier-1 Welford filter, the Tier-2 LangGraph Agent, hybrid Dual-RAG, and defensive security guardrails.

\section{Accuracy Evaluation \& McNemar's Statistical Test}
The accuracy evaluation reveals significant improvements in Precision, Recall, and F1-Score across the different configurations. Most notably, the holistic Configuration F achieves the optimal F1 balance by successfully eliminating the high volume of false positives generated by rule-based engines, while simultaneously enhancing the recall rate for Zero-day threats. The system's APT detection efficacy is demonstrated by the Tier-1 Welford filter capturing subtle anomalous signals, which then populate the agent's Threat Memory to expose multi-stage lateral movement chains. To prove the statistical significance of these accuracy improvements, McNemar's test was applied to compare the confusion matrices of the baseline Configuration B and the complete Configuration F. The resulting statistical calculation yielded a $p\text{-value} < 0.05$, thereby rejecting the null hypothesis and confirming that the performance gains are directly attributable to the architectural enhancements of SENTINEL rather than random chance.

\section{Performance Latency Evaluation \& Mann-Whitney U Test}
Performance evaluation focused on the end-to-end processing latency per network flow. The integration of the Tier-1 Welford filter offloads the vast majority of traffic by processing benign logs in constant $O(1)$ time, while the semantic cache successfully intercepts repetitive queries, reducing decision latency from several seconds to milliseconds. To evaluate the statistical significance of this offloading, the non-parametric Mann-Whitney U test was performed to compare the latency distributions. The analysis proved that the SENTINEL pipeline is significantly faster ($p\text{-value} < 0.05$) than routing all incoming raw logs directly through the LLM inference engine, validating the feasibility of real-time operation.

\section{Adversarial Robustness Evaluation (Security)}
To test the security posture of the cognitive layer, an adversarial AI manipulation dataset consisting of 45 specialized payloads was compiled. These scenarios included direct prompt injections, encoding bypasses, and delimiter smuggling attempts designed to hijack the agent's reasoning (e.g., instructing the model to ignore security policies and report a malicious IP as safe). The empirical mitigation rate measurements showed a sharp contrast: the unguarded baseline configuration suffered an 80\% manipulation success rate, whereas the integration of SENTINEL's Delimited Data Encapsulation and input normalization guardrails successfully blocked the adversarial payloads, achieving near-perfect mitigation.

\section{Integrity and RAG Quality Evaluation via LLM-as-a-Judge}
The final validation phase evaluated database integrity and retrieval quality. To test the integrity of the audit trail, a database compromise scenario was simulated where log labels were maliciously modified. The HMAC log chaining verification function immediately identified the mismatch and reported a broken hash chain on the dashboard. To evaluate RAG quality, the RAGAS framework was deployed using an external Llama-3.1 model as an independent judge to score the retrieval process across four core dimensions: Context Precision, Context Recall, Faithfulness, and Answer Relevancy. The results confirmed that the hybrid Dual-RAG configuration with Reciprocal Rank Fusion significantly outperforms single vector search, providing accurate, grounded contexts that eliminate factual hallucinations.

In summary, Chapter 4 verifies the technical prowess of SENTINEL through hard quantitative data and rigorous biostatistical tests. Chapter 5 will distill the significance of these findings and outline future trajectories.
