% =============================================================================
% FPT UNIVERSITY - MASTER OF SOFTWARE ENGINEERING (MSE) THESIS TEMPLATE
% Formatted strictly according to FPT MSE Thesis Guidelines (V5)
% =============================================================================
\documentclass[12pt,a4paper]{report}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T5]{fontenc} % Keep T5 for any occasional Vietnamese names
\usepackage[english]{babel}
\addto\captionsenglish{\renewcommand{\bibname}{References}}
\usepackage{times} % Times New Roman font
\usepackage{geometry}
\geometry{top=2.54cm, bottom=2.54cm, left=3.0cm, right=2.8cm} % Margin rules
\usepackage{graphicx}
\usepackage{setspace} % Double spacing package
\usepackage{fancyhdr} % Page numbering custom configuration
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{cite}
\usepackage{array}
\usepackage{longtable}
\usepackage{float}
\usepackage{listings}
\usepackage{xcolor}

% --- Continuously number figures and tables across chapters ---
\usepackage{chngcntr}
\counterwithout{figure}{chapter}
\counterwithout{table}{chapter}

% --- Custom caption label separator (period instead of colon) ---
\usepackage[labelsep=period]{caption}

% --- Custom List of Figures/Tables layout using tocloft ---
\usepackage{tocloft}
\renewcommand{\cftfigpresnum}{Figure~}
\renewcommand{\cftfigaftersnum}{.}
\renewcommand{\cfttabpresnum}{Table~}
\renewcommand{\cfttabaftersnum}{.}
\setlength{\cftfignumwidth}{2.3cm}
\setlength{\cfttabnumwidth}{2.0cm}

% Redefine list names to match screenshot (singular)
\renewcommand{\listtablename}{List of Table}
\renewcommand{\listfigurename}{List of Figure}

% --- Paragraph Formatting (Block Style: No indent, 6pt space above) ---
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}

% --- Headings Formatting (Maximum 3 levels) ---
\setcounter{secnumdepth}{3}
\titleformat{\chapter}[display]{\Large\bfseries\centering}{\chaptername\ \thechapter}{0.5em}{}
\titleformat{\section}{\large\bfseries}{\thesection}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}{\normalsize\itshape}{\thesubsubsection}{0.5em}{}

% --- Code Block Styling ---
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!5},
    keywordstyle=\color{blue},
    commentstyle=\color{green!60!black},
    stringstyle=\color{orange},
    showstringspaces=false,
    inputencoding=utf8,
    extendedchars=true
}

% --- Header/Footer Customization (Page number on Top-Right) ---
\pagestyle{fancy}
\setlength{\headheight}{15pt}
\fancyhf{}
\fancyhead[R]{\thepage} % Position page number on top right
\renewcommand{\headrulewidth}{0pt} % Remove header line

% Override standard plain style (which is automatically used on Chapter start pages)
\fancypagestyle{plain}{
    \fancyhf{}
    \fancyhead[R]{\thepage}
    \renewcommand{\headrulewidth}{0pt}
}

% =============================================================================
\renewcommand{\arraystretch}{1.3} % uniform vertical spacing for all tables
\begin{document}
\selectlanguage{english}

% -----------------------------------------------------------------------------
% FRONT MATTER (No page numbering on cover pages)
% -----------------------------------------------------------------------------
\pagestyle{empty}

% Cover Page 1
\begin{titlepage}
    \centering
    \vspace*{1cm}
    {\large\textbf{MINISTRY OF EDUCATION AND TRAINING}}\\[0.1cm]
    {\large\textbf{FPT UNIVERSITY}}\\[3cm]
    
    {\Large A TWO-TIER COGNITIVE ARCHITECTURE FOR AUTOMATED THREAT DETECTION AND RESPONSE USING AGENTIC AI}\\[3cm]
    
    by\\[1cm]
    
    Nguyen Duc Binh\\[3cm]
    
    A thesis submitted in conformity with the requirements\\
    for the degree of Master of Software Engineering\\[1cm]
    
    \vfill
    {\normalsize \copyright\ Copyright by Nguyen Duc Binh 2026}
\end{titlepage}

% Cover Page 2 (with Supervisors)
\begin{titlepage}
    \centering
    \vspace*{1cm}
    {\large\textbf{MINISTRY OF EDUCATION AND TRAINING}}\\[0.1cm]
    {\large\textbf{FPT UNIVERSITY}}\\[3cm]
    
    {\Large A TWO-TIER COGNITIVE ARCHITECTURE FOR AUTOMATED THREAT DETECTION AND RESPONSE USING AGENTIC AI}\\[3cm]
    
    by\\[1cm]
    
    Nguyen Duc Binh\\[3cm]
    
    A thesis submitted in conformity with the requirements\\
    for the degree of Master of Software Engineering\\[2cm]
    
    Supervisor:\\
    Dr. Bui Van Hieu\\
    Dr. Dang Van Hieu\\[1cm]
    
    \vfill
    {\normalsize \copyright\ Copyright by Nguyen Duc Binh 2026}
\end{titlepage}

% -----------------------------------------------------------------------------
% FRONT MATTER (ARABIC NUMERALS 1, 2, 3...)
% -----------------------------------------------------------------------------
\pagenumbering{arabic}
\pagestyle{fancy}
\setcounter{page}{1} % Starts from page 1 after cover pages

% Abstract Page
\newpage
\addcontentsline{toc}{chapter}{Abstract}
\begin{center}
    {\Large\textbf{Abstract}}
\end{center}
\vspace{0.5cm}

\begin{singlespace}
In the contemporary cybersecurity landscape, Security Operations Centers (SOCs) are consistently overwhelmed by the sheer volume of network alerts, leading to severe alert fatigue and the potential oversight of critical Advanced Persistent Threats (APTs). To address the inherent limitations of traditional, static rule-based systems and the latency bottlenecks of processing raw logs directly through Large Language Models (LLMs), this thesis introduces SENTINEL, a two-tier cognitive architecture for automated threat detection and response. The framework deploys Welford's online statistical algorithm at a stateless Transport Tier (Tier-1) to filter benign traffic at wire speed in $O(1)$ time and memory, while a LangGraph-driven Cognitive Agent Tier (Tier-2)—equipped with a Dual Retrieval-Augmented Generation (Dual-RAG) mechanism over MITRE ATT\&CK techniques and NIST SP 800-61r2 playbooks—performs grounded contextual reasoning behind cryptographic guardrails.

Evaluated on the CSE-CIC-IDS2018 and DAPT2020 datasets, SENTINEL escalates only the genuinely ambiguous minority of traffic to the LLM: the deterministic tier resolves a decision in roughly $0.6$\,ms versus a $4$--$6$\,s LLM inference, a separation confirmed as statistically significant by a Mann-Whitney~U test ($p = 2.8\times10^{-17}$). The online Welford filter caught all seven injected signature-less zero-day outliers that the static engine missed, and emergent multi-day correlation in the persistent Threat Memory recovered all three DAPT2020 APT campaigns (recall $= 1.0$). A cross-family LLM-as-Judge evaluation rated the agent's grounded reasoning at $3.91/5$ overall, with a Faithfulness of $4.00/5$, while a defense-in-depth guardrail design---Delimited Data Encapsulation, output sanitization, and HMAC log chaining---blocked all encoding-based injection payloads and routed residual semantic manipulations to human review while preserving forensic integrity. These findings demonstrate that pairing a deterministic wire-speed filter with an explainable, cryptographically isolated Agentic AI tier yields a scalable and security-hardened automation solution for modern network defense.


\end{singlespace}

% Acknowledgements Page
\newpage
\addcontentsline{toc}{chapter}{Acknowledgments}
\begin{center}
    {\Large\textbf{Acknowledgments}}
\end{center}
\vspace{0.5cm}

I would like to express my deepest and most sincere gratitude to my supervisors, Dr. Bui Van Hieu and Dr. Dang Van Hieu. Their precise academic direction, sharp professional feedback, and continuous guidance throughout this research have served as the cornerstone for the successful completion of this thesis.

I am also profoundly grateful to all the faculty members and staff at FPT University, particularly the Master of Software Engineering (MSE) program, for equipping me with invaluable academic knowledge, fostering a highly professional and modern research environment, and providing excellent resources during my master's studies.

Furthermore, I would like to extend my appreciation to the management and my colleagues at FPT Software for their support, understanding, and valuable professional discussions, which greatly helped me balance my professional duties and academic pursuits.

Finally and most importantly, I dedicate a special acknowledgment to my loving family, especially my wife and young child. Their silent sacrifices, patience, understanding, and unwavering moral support during my intensive experimental runs provided the ultimate strength and motivation for me throughout this master's journey.


% Lists and Table of Contents
\newpage
\listoffigures
\newpage
\listoftables
\newpage
\addcontentsline{toc}{chapter}{List of Abbreviations}
\begin{center}
    {\Large\textbf{List of Abbreviations}}
\end{center}
\vspace{0.5cm}

\begin{longtable}{lp{0.72\textwidth}}
\textbf{Abbreviation} & \textbf{Full Description} \\
\midrule
AI & Artificial Intelligence \\
APT & Advanced Persistent Threat \\
C2 & Command and Control \\
CNN & Convolutional Neural Network \\
CPU & Central Processing Unit \\
CSV & Comma-Separated Values \\
CTI & Cyber Threat Intelligence \\
CVE & Common Vulnerabilities and Exposures \\
DAG & Directed Acyclic Graph \\
DAPT & Dynamic/Delayed Advanced Persistent Threat \\
DL & Deep Learning \\
DoS & Denial of Service \\
EDR & Endpoint Detection and Response \\
FAISS & Facebook AI Similarity Search \\
FSM & Finite State Machine \\
GGUF & GPT-Generated Unified Format \\
GPU & Graphics Processing Unit \\
HITL & Human-in-the-Loop \\
HMAC & Hash-based Message Authentication Code \\
IDS & Intrusion Detection System \\
IOC & Indicator of Compromise \\
IP & Internet Protocol \\
IPS & Intrusion Prevention System \\
JSON & JavaScript Object Notation \\
LLM & Large Language Model \\
LRU & Least Recently Used \\
LSTM & Long Short-Term Memory \\
MAS & Multi-Agent System \\
ML & Machine Learning \\
NGFW & Next-Generation Firewall \\
NIST & National Institute of Standards and Technology \\
OWASP & Open Web Application Security Project \\
PCAP & Packet Capture \\
RAG & Retrieval-Augmented Generation \\
RAGAS & Retrieval Augmented Generation Assessment \\
RAM & Random Access Memory \\
RBA & Runbook Automation \\
RCE & Remote Code Execution \\
RRF & Reciprocal Rank Fusion \\
SHA & Secure Hash Algorithm \\
SIEM & Security Information and Event Management \\
SOC & Security Operations Center \\
SQLi & SQL Injection \\
SSTI & Server-Side Template Injection \\
SVM & Support Vector Machine \\
TTP & Tactics, Techniques, and Procedures \\
TTL & Time to Live \\
URL & Uniform Resource Locator \\
VRAM & Video Random Access Memory \\
WAF & Web Application Firewall \\
XSS & Cross-Site Scripting \\
\bottomrule
\end{longtable}
\newpage
\tableofcontents

% -----------------------------------------------------------------------------
% MAIN BODY
% -----------------------------------------------------------------------------
\newpage
\onehalfspacing % Apply 1.5 Spacing rules to main body text

% Include modular chapter files
\setcounter{table}{0}
\setcounter{figure}{0}
\include{chapters/ch1_introduction}
\include{chapters/ch2_theoretical_background}
\include{chapters/ch3_system_design}
\include{chapters/ch4_experiments_evaluation}
\include{chapters/ch5_conclusion}

% -----------------------------------------------------------------------------
% REFERENCES (Continuous page numbering)
% -----------------------------------------------------------------------------
\newpage
\singlespacing % References are typically single-spaced

\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{References}

\bibitem{ids2018}
I. Sharafaldin, A. H. Lashkari, and A. A. Ghorbani, ``Toward Generating a New Intrusion Detection Dataset and Intrusion Traffic Characterization,'' in \textit{Proceedings of the 4th International Conference on Information Systems Security and Privacy (ICISSP)}, 2018, pp. 108--116.

\bibitem{siem2021}
G. González-Granadillo, S. González-Zarzosa, and R. Diaz, ``Security Information and Event Management (SIEM): Analysis, Trends, and Usage in Critical Infrastructures,'' \textit{Sensors}, vol. 21, no. 14, art. 4759, 2021.

\bibitem{bigdata2019}
R. Zuech, T. M. Khoshgoftaar, and R. Wald, ``Intrusion Detection and Big Heterogeneous Data: A Survey,'' \textit{Journal of Big Data}, vol. 2, art. 3, 2015.

\bibitem{apt_stealth2021}
M. Alshamrani \textit{et al.}, ``A Survey on Advanced Persistent Threats: Techniques, Solutions, Challenges, and Research Opportunities,'' \textit{IEEE Communications Surveys \& Tutorials}, vol. 21, no. 2, pp. 1851--1877, 2019.

\bibitem{alertfatigue2022}
S. Tariq, M. Baruwal Chhetri, S. Nepal, and C. Paris, ``Alert Fatigue in Security Operations Centres: Research Challenges and Opportunities,'' \textit{ACM Computing Surveys}, vol. 57, no. 9, art. 224, 2025.

\bibitem{welford1962}
B. P. Welford, ``Note on a Method for Calculating Corrected Sums of Squares and Products,'' \textit{Technometrics}, vol. 4, no. 3, pp. 419--420, 1962.

\bibitem{rag2020}
P. Lewis \textit{et al.}, ``Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks,'' \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 33, pp. 9459--9474, 2020.

\bibitem{langgraph2024}
LangChain AI, ``LangGraph: Multi-Actor, Stateful LLM Applications,'' \textit{GitHub repository}, 2024. [Online]. Available: https://github.com/langchain-ai/langgraph

\bibitem{promptinjection2023}
K. Greshake, S. Abdelnabi, S. Mishra, A. Endres, T. Holz, and M. Fritz, ``More than you've asked for: A Comprehensive Analysis of Novel Prompt Injection Threats to Application-Integrated Large Language Models,'' \textit{arXiv preprint arXiv:2302.12173}, 2023.

\bibitem{nist80061}
P. Cichonski, T. Millar, T. Grance, and K. Scarfone, ``Computer Security Incident Handling Guide,'' NIST Special Publication 800-61 Revision 2, 2012.

\bibitem{rrf2009}
G. V. Cormack, C. L. A. Clarke, and S. Buettcher, ``Reciprocal Rank Fusion Outperforms Condorcet and Individual Bootstrap Product Algorithms,'' in \textit{Proceedings of the 32nd International ACM SIGIR Conference on Research and Development in Information Retrieval}, 2009.

\bibitem{watchtower2026}
R. Pandey and A. Bhujang, ``Poisoning the Watchtower: Prompt Injection Attacks Against LLM-Augmented Security Operations Through Adversarial Log Content,'' \textit{arXiv preprint arXiv:2605.24421}, 2026.

\bibitem{cyberrag2025}
F. Blefari, C. Cosentino, F. A. Pironti, A. Furfaro, and F. Marozzo, ``CyberRAG: An Agentic RAG cyber attack classification and reporting tool,'' \textit{Future Generation Computer Systems}, vol. 176, art. 108186, 2026.

\bibitem{lang2026}
A. Abdennebi, N. Kara, L. Lahlou, and H. Ould-Slimane, ``LanG -- A Governance-Aware Agentic AI Platform for Unified Security Operations,'' \textit{arXiv preprint arXiv:2604.05440}, 2026.

\bibitem{splunkllm2026}
R. Sahay \textit{et al.}, ``Policy-Guided Threat Hunting: An LLM Enabled Framework with Splunk SOC Triage,'' \textit{arXiv preprint arXiv:2603.23966}, 2026.

\bibitem{vaswani2017}
A. Vaswani \textit{et al.}, ``Attention is All You Need,'' in \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 30, pp. 5998--6008, 2017.

\bibitem{quantization2021}
A. Gholami \textit{et al.}, ``A Survey of Quantization Methods for Efficient Neural Network Inference,'' \textit{arXiv preprint arXiv:2103.13630}, 2021.

\bibitem{faiss2019}
J. Johnson, M. Douze, and H. Jégou, ``Billion-Scale Similarity Search with GPUs,'' \textit{IEEE Transactions on Big Data}, vol. 7, no. 3, pp. 535--547, 2019.

\bibitem{bm25_2009}
S. Robertson and H. Zaragoza, ``The Probabilistic Relevance Framework: BM25 and Beyond,'' \textit{Foundations and Trends in Information Retrieval}, vol. 3, no. 4, pp. 333--380, 2009.

\bibitem{he2017drain}
P. He, J. Zhu, S. He, J. Li, and M. R. Lyu, ``Drain: An Online Log Parsing Approach with Fixed Depth Tree,'' in \textit{Proceedings of the IEEE International Conference on Web Services (ICWS)}, 2017, pp. 33--40.

\bibitem{gemma_2_2024}
Gemma Team, Google, ``Gemma 2: Improving Open Language Models at a Practical Size,'' \textit{arXiv preprint arXiv:2408.00118}, 2024.

\bibitem{strom2018mitre}
B. E. Strom, A. Applebaum, D. P. Miller, K. C. Nickels, A. G. Pennington, and C. B. Thomas, ``MITRE ATT\&CK: Design and Philosophy,'' The MITRE Corporation, Technical Report MP180360, 2018.

\bibitem{reimers2019sentence}
N. Reimers and I. Gurevych, ``Sentence-BERT: Sentence Embeddings using Siamese BERT-Networks,'' in \textit{Proceedings of the Conference on Empirical Methods in Natural Language Processing (EMNLP)}, 2019.

\bibitem{zheng2023judge}
L. Zheng \textit{et al.}, ``Judging LLM-as-a-Judge with MT-Bench and Chatbot Arena,'' in \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 36, 2023.

\bibitem{dapt2020}
S. Myneni, A. Chowdhary, A. Sabur, S. Sengupta, G. Agrawal, D. Huang, and M. Kang, ``DAPT 2020 -- Constructing a Benchmark Dataset for Advanced Persistent Threats,'' in \textit{Deployable Machine Learning for Security Defense (MLHat 2020)}, Communications in Computer and Information Science, vol. 1271, Springer, 2020, pp. 138--163.

\bibitem{ragas2023}
S. Es, J. James, L. Espinosa-Anke, and S. Schockaert, ``RAGAS: Automated Evaluation of Retrieval Augmented Generation,'' \textit{arXiv preprint arXiv:2309.15217}, 2023.

\bibitem{mcnemar1947}
Q. McNemar, ``Note on the Sampling Error of the Difference Between Correlated Proportions or Percentages,'' \textit{Psychometrika}, vol. 12, no. 2, pp. 153--157, 1947.

\bibitem{mannwhitney1947}
H. B. Mann and D. R. Whitney, ``On a Test of Whether One of Two Random Variables is Stochastically Larger than the Other,'' \textit{Annals of Mathematical Statistics}, vol. 18, no. 1, pp. 50--60, 1947.

\bibitem{hmac1997}
H. Krawczyk, M. Bellare, and R. Canetti, ``HMAC: Keyed-Hashing for Message Authentication,'' Internet Engineering Task Force, RFC 2104, 1997.

\bibitem{owasp_llm2025}
OWASP GenAI Security Project, ``OWASP Top 10 for Large Language Model Applications,'' Version 2025, The OWASP Foundation, 2025. [Online]. Available: https://genai.owasp.org/

\bibitem{nist80207}
S. Rose, O. Borchert, S. Mitchell, and S. Connelly, ``Zero Trust Architecture,'' NIST Special Publication 800-207, National Institute of Standards and Technology, 2020.

\end{thebibliography}

\end{document}
